Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-05 paper/main.tex
Updates

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association.

@@

\subsubsection{Both $ X $ and $ Y $ are continuous}
When both $ X $ and $ Y $ are continuous variables, Pearson's correlation
coefficient is typically used. In the case when $ \bm{Z} = \emptyset $, the
correlation coefficient is given as:

@@

where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively. When $ \bm{Z} \neq \emptyset $, partial correlation coefficient
can be used. One of the ways to estimate it is to use two regression
models $ E_X: X \sim \bm{Z} $, $ E_Y: Y \sim \bm{Z} $, and compute the Pearson's 
correlation coefficient between the residuals of these regression models:

@@

\paragraph{$ X $ is ordinal and $ Y $ is continuous or ordinal}
When one of the variables is ordinal and the other is continuous polyserial
correlation can be used. When both the variables are ordinal polychoric
correlation can be used. Both polychoric and polyserial correlation make the
assumption that the ordinal variable has an underlying normal distributed
continuous variable, and the ordinal variable has been generated by
thresholding this latent variable. The method then tries to maximize the
likelihood of data given polychoric or polyserial correlation coefficient. This
method gives us values that are Pearson's correlation coefficient.

>>>>

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.

@@

\subsubsection{Both $ X $ and $ Y $ are continuous}
When both $ X $ and $ Y $ are continuous variables, Pearson's correlation
coefficient is typically used. In the case when $ \bm{Z} = \emptyset $, the
correlation coefficient is defined as:

@@

When $ \bm{Z} \neq \emptyset $, partial correlation coefficient
can be used. One of the ways to estimate it is to use two regression
models $ E_X: X \sim \bm{Z} $, $ E_Y: Y \sim \bm{Z} $, and compute the Pearson's 
correlation coefficient between the residuals of these regression models:

@@

\paragraph{$ X $ is ordinal and $ Y $ is continuous or ordinal}
When one of the variables is ordinal and the other is continuous polyserial
correlation can be used. When both the variables are ordinal polychoric
correlation can be used. Both polychoric and polyserial correlation make the
assumption that the ordinal variable has an underlying normal distributed
continuous variable, and the ordinal variable has been generated by
thresholding this latent variable. The method then tries to maximize the
likelihood of data given polychoric or polyserial correlation coefficient. This 
method gives us the covariance matrix, $\Sigma$ on the variables $ X $, $ Y $, and $ \bm{Z} $.

\begin{equation}
	\rho_{X, Y} = \Sigma_{XY}
\end{equation}

When $ \bm{Z} \neq \emptyset $,

\begin{equation}
	\rho_{X, Y; \bm{Z}} = - \frac{\Sigma^{-1}_{XY}}{\sqrt{\Sigma^{-1}_{XX} \Sigma^{-1}_{YY}}}
\end{equation}
